---
title: "hoodscanR & scider: Spatial downstream analysis workshop"
author:
  - name: Ning Liu
    affiliation:
      - SAiGENCI, University of Adelaide, Adelaide, SA 5000, Australia
      - Adelaide Centre of Epigenetics, School of Biomedicine, University of Adelaide, Adelaide, SA 5000, Australia
    email: ning.liu@adelaide.edu.au
  - name: Nora Liu
    affiliation:
      - SAiGENCI, University of Adelaide, Adelaide, SA 5000, Australia
      - Adelaide Centre of Epigenetics, School of Biomedicine, University of Adelaide, Adelaide, SA 5000, Australia
    email: nora.liu@adelaide.edu.au
  - name: Monika Mohenska
    affiliation:
      - SAiGENCI, University of Adelaide, Adelaide, SA 5000, Australia
      - Adelaide Centre of Epigenetics, School of Biomedicine, University of Adelaide, Adelaide, SA 5000, Australia
    email: monika.mohenska@adelaide.edu.au
  - name: Jose Polo
    affiliation:
      - SAiGENCI, University of Adelaide, Adelaide, SA 5000, Australia
      - Adelaide Centre of Epigenetics, School of Biomedicine, University of Adelaide, Adelaide, SA 5000, Australia
      - Australian Regenerative Medicine Institute and Systems Biology Institute Australia, Monash University, Clayton, VIC 3800, Australia
    email: jose.polo@adelaide.edu.au
date: "`r format(Sys.time(), '%b %Y')`"
output:
  prettydoc::html_pretty:
    theme: cayman
    toc: yes
    toc_depth: 2
    number_sections: yes
    fig_caption: yes
    df_print: paged
nocite: | 
  @R-ggplot2, @R-knitr, @R-BiocWorkflowTools, @R-rmarkdown, @R-prettydoc
vignette: >
  %\VignetteIndexEntry{hoodscanR & scider: Spatial downstream analysis workshop}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
keywords: spatial transcriptomics, neighbourhood, density
---

<p>

**R version**: `r R.version.string` <br /> **Bioconductor version**: `r BiocManager::version()` 

</p>

# Welcome

This 90‑minute, hands‑on workshop introduces two complementary Bioconductor packages for spatial transcriptomics (ST):

- **hoodscanR** — discover *local cell neighborhoods* and perform neighborhood‑based downstream analysis.  
- **scider** — estimate *cell type–specific spatial densities*, identify cell-type-specific ROIs and contours, and run density‑driven analyses.

We’ll start with **local** structure (hoodscanR), then **zoom out** to density (scider).

> **Data**  
> Use either: (a) a published Xenium example in the vignettes below, or (b) your own ST dataset (counts + coordinates).

## Getting started

```{r}
#install.packages("BiocManager")

# If needed:
#BiocManager::install(c("hoodscanR", "scider", "tidyverse", 
#                       "patchwork", "ggrepel", "Seurat", "edgeR",
#                        "zellkonverter"))

suppressPackageStartupMessages({
  library(hoodscanR)
  library(scider)
  library(tidyverse)
})
```

# Part A — Local neighborhoods with hoodscanR

Below is an edited version of the **hoodscanR** quick start, adapted for a live workshop.

```{=html}
<style type="text/css">
  body { font-size: 14pt; }
</style>
```

```{r}
knitr::opts_chunk$set(echo = TRUE, message = FALSE, warning = FALSE, dpi = 96)
```

## Introduction

**hoodscanR** is an R package for analysing *cellular neighborhoods* in single‑cell‑resolution ST data.  
It is built on the `r BiocStyle::Biocpkg("SpatialExperiment")` infrastructure and produces per‑cell neighborhood annotations, co‑localisation statistics, and neighborhood‑based clustering.

## Installation

```{r}
#| eval: false
if (!require("BiocManager", quietly = TRUE)) install.packages("BiocManager")
BiocManager::install("hoodscanR")

# Development version
# devtools::install_github("DavisLaboratory/hoodscanR")
```

```{r}
suppressPackageStartupMessages({
  library(SummarizedExperiment)
  library(SingleCellExperiment)
  library(SpatialExperiment)
})
library(hoodscanR)

ws_colors <- c(
  "#FFBE7D", "#8E3B5D", "#1B263B", "#778DA9", "#BB3E03",
  "#D37295", "#0A9396", "#8CD17D", "#FFD700", "#90EE90",
  "#9B2226", "#E15759", "#9467BD", "#F28E2B", "#2EFEF7",
  "#EE9B00", "#CA6702", "#94D2BD", "#E9D8A6", "#005F73",
  "#32CD32", "#C5B0D5", "#3B3B3B", "#000080", "#2F4F4F",
  "#FF00FF", "#7BA05B", "#AFEEEE", "#FF9D9A", "#BAB0AC",
  "#59A14F"
)
```

## Data exploration

In this workshop we use a breast‑cancer tissue section from *Janesick et al.* (2023, *Nature Communications*), which introduced the Xenium in‑situ platform. This FFPE human breast tumour (ER+/HER2+, Stage II‑B) was profiled with a targeted 313‑gene panel using rolling‑circle amplification and cyclic fluorescent probe imaging, providing sub‑cellular transcript localisation. This dataset is ideal for combining neighborhood‑level and density‑level spatial analyses.

```{r}
spe <- readRDS("../inst/extdata/spe_S1_Rep1.RDS")
```

Let’s confirm the structure of the `SpatialExperiment` object and its key elements.

```{r}
spe          # dimensions, assays, and metadata
assayNames(spe)
head(colData(spe))
table(colData(spe)$cell_type)
head(spatialCoords(spe))
```

The `readHoodData` function can format the spatialExperiment input object 
as desired for all other functions in the `hoodscanR` package.


```{r}
spe <- readHoodData(spe, anno_col = "cell_type")
```

We can have a look at the tissue and cell positions by using the function `plotTissue`.


The `plotTissue` function is very flexible, it takes all parameters that can go
into `geom_point()` or `aes()`. And the output is a ggplot object, therefore is
easily customised.

```{r}
#| fig-width: 10
#| fig-height: 6
plotTissue(spe, color = cell_annotation, size = 0.1, alpha = 0.9) +
  scale_color_manual(values = ws_colors) +
  guides(color = guide_legend(override.aes = list(shape = 15, size = 5)))
```

## Neighborhood detection

Identify the *k* nearest cells for each cell (reduced to `k = 50` for speed during the workshop):

```{r}
fnc <- findNearCells(spe, k = 50)
lapply(fnc, function(x) x[1:10, 1:5])  # peek at annotation/distance matrices
```

Compute per‑cell neighborhood probabilities.
The `scanHoods` function contains the core algorithm of hoodscanR, which is controlled
by the hyperparameter `tau`.
By default, `tau` is set to median(m**2)/5 as `m` is the distance matrix (this was tested in multiple datasets).
Smaller `tau` value assign greater weights to nearby cells, while larger value consider more distant cells as contributors to the neighborhood.
In extreme case, if one want to have equal weights of all k cells, set parameter `mode` to `smoothFadeout`, it will search for the desired `tau` automatically.

```{r}
pm <- scanHoods(fnc$distance)
pm[1:10, 1:5]
```

And the the probablities add up to 1.

```{r}
rowSums(pm)[sample(1:nrow(pm),50)]
```


The probability matrix can then be aggregated by annotation (cell type in this case):

```{r}
hoods <- mergeByGroup(pm, fnc$cells)
hoods[1:10, ]
```

## neighborhood composition and clustering

We plot randomly plot 10 cells to see the output of neighborhood scanning 
using `plotHoodMat`. In this plot, each value represent the probability of the each
cell (each row) located in each cell type neighborhood. 

```{r}
plotHoodMat(hoods, n = 10, hm_height = 5)
```

Or to check the cells-of-interest with the parameter `targetCells`.

```{r}
plotHoodMat(hoods, targetCells = c("900", "100000"), hm_height = 1)
```

We can then attach neighborhood probabilities back to `spe` and compute summary metrics.

We can use `calcMetrics` to calculate entropy and perplexity of the probability
matrix so that we can have a summarisation of the neighborhood distribution 
across the tissue slide, i.e. where neighborhood is more distinct and where is 
more mixed.

While both entropy and perplexity measure the mixture of neighborhood
of each cell, perplexity can be more intuitive for the human mind as the value
of it actually mean something. For example, perplexity of 1 means the cell is
located in a very distinct neighborhood, perplexity of 2 means the cell is
located in a mixed neighborhood, and the probability is about 50% to 50%.

```{r}
spe <- mergeHoodSpe(spe, hoods)
spe <- calcMetrics(spe, pm_cols = colnames(hoods))  # entropy & perplexity
```

Map entropy/perplexity across the tissue:

```{r}
#| fig-width: 10
#| fig-height: 6
plotTissue(spe, color = entropy, size = 0.1) +
  scale_color_gradient(low = "darkslateblue", high = "gold")
```


```{r}
#| fig-width: 10
#| fig-height: 6
plotTissue(spe, color = perplexity, size = 0.1) +
  scale_color_gradient(low = "darkslategray", high = "gold")
```

We can perform neighborhood colocalisation analysis using `plotColocal`.
This function compute pearson correlation on the probability distribution of each cell.
Here we can see in the test data, immune cells are more likely to co-localise, luminal progeniotr (LP) cells and Basal cells are more likely to colocalise. Cancer cells have overall negative correlation to other immune cells, indicating no immune infiltration on this sample.

```{r}
hood_colocal_p <- plotColocal(spe, pm_cols = colnames(hoods))
```

We can cluster the cells by their neighborhood probability distribution using `clustByHood`.


```{r}
set.seed(88)

spe <- clustByHood(spe, pm_cols = colnames(hoods), k = 9)
```

We can see what are the neighborhood distributions look like in each cluster using `plotProbDist`.

```{r}
#| fig-width: 10
#| fig-height: 7
plotProbDist(
  spe, pm_cols = colnames(hoods), by_cluster = TRUE, plot_all = TRUE,
  fill = hoods, show_clusters = as.character(seq(9)), outlier.size = 0.1
) + 
  scale_fill_manual(values = ws_colors) + 
  theme(legend.position = "none")
```

We can plot the clusters on the tissue slide, agian using `plotTissue`.

```{r}
#| fig-width: 10
#| fig-height: 6
plotTissue(spe, color = clusters, size = 0.1) +
  scale_color_manual(values = ws_colors[10:20]) +
  guides(color = guide_legend(override.aes = list(shape = 15, size = 5)))
```

We can then mannual annotated the cluster by the distribution of
cell type neighborhoods.


```{r}
clust_anno <- data.frame(
  clusters = as.character(seq(9)),
  cluster_annotation = c(
    "NearCancer","Cancer","ImmuneTME",
    "Cancer","ImmuneInfiltration","ImmuneInfiltration",
    "Cancer","TME","ImmuneTME"
  )
)

spe$cluster_annotation <- as.data.frame(colData(spe)) %>%
  #dplyr::select(-cluster_annotation) %>%
  dplyr::left_join(clust_anno, by = "clusters") %>%
  dplyr::pull(cluster_annotation)

```


```{r}
#| fig-width: 10
#| fig-height: 6
plotTissue(spe, color = cluster_annotation, size = 0.1, reverseY = FALSE) +
  scale_color_manual(values = ws_colors[3:10]) +
  guides(color = guide_legend(override.aes = list(shape = 15, size = 5)))
```

## Comparing expression across neighborhood contexts

As an illustration, we want to look at if spatial neighborhoods play a role
in cancer cells from "Near Cancer" and "ImmuneInfiltration" neighborhoods.

We can first subset cells that are "Cancer" cell types and located in "NearCancer" and "ImmnueInfiltration" neighborhoods.

```{r}
spe_c_nc <- spe[, spe$cluster_annotation == "NearCancer" & spe$cell_annotation == "Cancer"]
spe_c_nc$group <- "Cancer_NearCancerHoods"

spe_c_ii <- spe[, spe$cluster_annotation == "ImmuneInfiltration" & spe$cell_annotation == "Cancer"]
spe_c_ii$group <- "Cancer_ImmuneInfiltrationHoods"

spe_subset <- cbind(spe_c_nc, spe_c_ii)
```

```{r}
#| fig-width: 10
#| fig-height: 6
plotTissue(spe_subset, size = 0.1, color = group, reverseY = FALSE) +
  scale_color_manual(values = c("tomato", "royalblue")) +
  guides(color = guide_legend(override.aes = list(shape = 15, size = 5)))
```
We can then perform differential expression (DE) analysis comparing cancer cells from different neighborhoods.
DE can be performed in single-cell style, i.e. using tools like `Seurat` or `Scran`.
Here we demonstrate a quick one using the `FindMarkers` function from `Seurat`.

When data available (data with replicates), we strongly recommend using pseudo-bulk strategy with DE pipeline such as `edgeR` or `limma-voom` to give a more robust statistical inference.

```{r}
library(Seurat)

seu <- CreateSeuratObject(
  counts = assay(spe_subset, "counts"),
  meta.data = as.data.frame(colData(spe_subset))
) %>% NormalizeData() %>% ScaleData()

Idents(seu) <- "group"
c.de <- FindMarkers(seu, ident.1 = "Cancer_NearCancerHoods",
                         ident.2 = "Cancer_ImmuneInfiltrationHoods",
                         verbose = FALSE)

de_df <- c.de %>%
  dplyr::mutate(score = avg_log2FC * -log10(p_val_adj + 8.00000e-324) * (pct.1 - pct.2)) %>%
  dplyr::arrange(dplyr::desc(score))

Seurat::DoHeatmap(seu, features = rownames(head(de_df, n = 30)), label = FALSE)
```

> **Try it yourself:** Re‑run the analysis with `k = 30` and `k = 80` when running `findNearCells`. How do entropy/perplexity or neighborhoods change?


# Part B — Cell‑type density, ROI and contours identification with scider

```{=html}
<style type="text/css">
  body { font-size: 12pt; }
</style>
```

```{r}
knitr::opts_chunk$set(echo = TRUE, message = FALSE, warning = FALSE, dpi = 200)
suppressWarnings(library(tidyverse))
suppressMessages(library(scider))
suppressMessages(library(SpatialExperiment))
```

**scider** models *global cell‑type density* within a tissue section using the `SpatialExperiment` class, and supports co‑localisation, ROI detection, and differential density analyses.

## Installation

```{r}
#| eval: false
if (!require("BiocManager", quietly = TRUE)) install.packages("BiocManager")
BiocManager::install("scider")

# Development version
# devtools::install_github("ChenLaboratory/scider")
```

```{r}
library(scider)
```

scider is a R package providing functions to model the global density of cells
in a slide of spatial transcriptomics data. 
Same as hoodscanR, all functions in the package are built based on the
SpatialExperiment object, allowing integration into various
ST-related packages from Bioconductor.
After modelling density, the package allows for serveral
downstream analysis, including colocalization analysis,
boundary detection analysis and differential density analysis.

## Load data

Again, we will use the same Xenium Breast Cancer dataset.

```{r}
spe <- readRDS("../inst/extdata/spe_S1_Rep1.RDS")
```

Similarly, in scider, we can use the function `plotSpatial` to
visualise the cell position and color the cells by cell types. 

```{r}
plotSpatial(spe, group.by = "cell_type", pt.alpha = 0.8)
```

## Grid‑based analysis

### Density calculation

We can perform density calculation for each cell type using function `gridDensity`.
The calculated density and grid information are saved in the metadata of the
SpatialExperimnet object.

```{r}
spe <- gridDensity(spe, id = "cell_type")
names(metadata(spe))
metadata(spe)$grid_density
```

We can visualise the overall cell density of the whole tissue using function `plotDensity`.

```{r}
plotDensity(spe)
```

The parameter `probs` control the filtering when plotting.
Since we didn't specify cell type of interest in `coi` parameter when running `gridDensity`.
It will compute cell density of all cell types.
In a tumor sample, usually cancer region would have higher cell density.

```{r}
plotDensity(spe, probs = 0)
```

We can also visualise the density of individual cell type, e.g., fibroblast cells or cancer cells.

```{r}
#| fig-width: 12
#| fig-height: 6
plotDensity(spe, coi = "Fibroblast") + plotDensity(spe, coi = "Cancer")
```

### Find Regions of Interest (ROIs)

After obtaining grid-based density for each COI, we can then detect
regions-of-interest (ROIs) based on density or select by user.

To detect ROIs automatically, we can use the function `findROI`.

The detected ROIs are saved in the metadata of the SpatialExperiment object.

Here we identify ROIs based on the cancer cell density.

```{r}
spe <- findROI(spe, coi = "Cancer")
metadata(spe)$cancer_roi
```

We can visualise the ROIs with function `plotROI`.

```{r}
plotROI(spe, roi = "Cancer")
```

Many parameters can be tuned in `findROI()`.
The `method` control what algorithm to be used when defining ROIs
(Greedy, Walktrap, Connected, HDbscan etc).
The `probs` control the proportion of cells (based on density) that 
go into the identification of ROIs.

```{r}
#| fig-width: 12
#| fig-height: 10
spe1 <- findROI(spe, coi = "Cancer", probs = 0.6)
spe2 <- findROI(spe, coi = "Cancer", probs = 0.6, method = "connected")
spe3 <- findROI(spe, coi = "Cancer", probs = 0.6, method = "walktrap")
spe4 <- findROI(spe, coi = "Cancer", probs = 0.6, method = "hdbscan")

(plotROI(spe1, roi = "Cancer") + ggtitle("Greedy")) + 
  (plotROI(spe2, roi = "Cancer") + ggtitle("Connected")) +
  (plotROI(spe3, roi = "Cancer") + ggtitle("Walktrap")) +
  (plotROI(spe4, roi = "Cancer") + ggtitle("HDbscan"))
```

```{r}
spe <- findROI(spe, coi = "Cancer", probs = 0.6)
```


### Testing relationships between cell types

After defining ROIs, we can then test the relationship between any two
cell types within each ROI or overall but account for ROI variation using
a cubic spline or a linear fit. This can be done with function `corrDensity`.

```{r}
cor.res <- corDensity(spe, roi = "Cancer")
```

We can see the correlation between each pair of cell types in each ROI.

```{r}
cor.res$ROI
```

Or the correlation between each pair of cell types across the whole slide:

```{r}
cor.res$overall
```

We can then visualise the statistics between each pair of cell types using
function `plotCorHeatmap` in the ROIs

```{r}
#| fig-width: 6
#| fig-height: 12
plotCorHeatmap(cor.res$ROI)
```

Or the correlation between cell type pairs across the whole slide:
```{r}
#| fig-width: 5
#| fig-height: 5
plotCorHeatmap(cor.res$overall)
```

We can look back at the co-localisation done by hoodscanR, since we were using 
cell types as the annotation to calculate the neighborhood probability matrix,
we can see that at whole slide level, both methods generate very similar results.

```{r}
hood_colocal_p
```

But in scider, we can focus on ROIs, we can also visualise a fitted 
pairwise relationship and its heterogeneity across ROIs.

We can see that in different cancer ROIs, the co-localisation status between 
basal cells and luminal progenitor are different. For example, they are positively
correlated in ROI1 and ROI4, but negatively correlated in ROI12 and ROI16.

Spatial pattern might be vary depends on tissue context, using scider can offer
a more precise approach by examining it in different ROI.

```{r}
#| fig-width: 12
#| fig-height: 6
plotDensCor(spe, celltype1 = "Basal", celltype2 = "LP", roi = "Cancer") + plotROI(spe, roi = "Cancer")
```


## Contour‑based analysis

Based on the grid density, we can ask many biological question about the data.
For example, we would like to know if a certain cell type that are located in
high density of cancer cells are different to the same cell type from
a different level of cancer region.


### Cell annotation by contour level

To address this question, we first need to divide cells into different
levels of grid density. This can be done using a contour identification
strategy with function `getContour`.

```{r}
spe <- getContour(spe, coi = "Cancer")
```

Different level of contour can be visualised with cells using `plotContour`.

```{r}
plotContour(spe, coi = "Cancer")
```


We can then annotate cells by their locations within each contour
using function `allocateCells`.


```{r}
spe <- allocateCells(spe)
```

We can visualise the continous regions on the tissue based on cancer density.

```{r}
plotSpatial(spe, group.by = "cancer_contour", pt.alpha = 0.5)
```

Compare composition per level:

```{r}
plotCellCompo(spe, contour = "Cancer")
plotCellCompo(spe, contour = "Cancer", self.included = FALSE)
```

## Pseudobulk

To investigate if cell types from different cancer contour regions have different
transcriptomics profiles, one way to do this is by using the `spe2PB` function
to pseudobulk the data by contour and cell types.


```{r}
y <- spe2PB(spe, contour = "Cancer")

y
dim(y)
```

Now that the object become a DGEList object, we can then use package such as 
`edgeR` to examine the data.

We can see that in PCA space, the pseudobulk data are clustered by cell type,
and within cell type, we can observe a trend of variance by contour level.

```{r}
#| fig-width: 9
#| fig-height: 7
y <- edgeR::calcNormFactors(y)
pca.res <- prcomp(t(edgeR::cpm(y, log = TRUE)), rank. = 10)$x

cbind(y$samples, pca.res) %>%
  ggplot(aes(PC1, PC2, color = cell_type)) +
  geom_point(size = 5) +
  scale_color_manual(values = ws_colors) +
  ggrepel::geom_text_repel(aes(label = contour.level)) +
  theme_classic()
```



# Part C — Useful utilities for Python tools

Many tools aimming to analyse ST data have built in Python, especially tools that 
built based on deep-learning algorithms.
In python, most ST data are handled using the `AnnData` infrastructure.
Therefore, it's important to learn how to transfer SpatialExperiment object data
into an `AnnData` object in Python. Which can then be used for many
Python-based methods.

In `AnnData` object, the spatial coordinates is usually store in the `obsm` slot,
which is used to store all the dimension reduction results. So we need to first
transfer the cell coordinates to the `reducedDims` slot of the object.

Then we can output a `.h5ad` file from the SpatialExperiment object, which can 
then read by `scanpy.read_h5ad` via the `scanpy` library in Python.


```{r}
spe <- readRDS("../inst/extdata/spe_S1_Rep1.RDS")
```


```{r}
reducedDims(spe) <- list(spatial = spatialCoords(spe))
zellkonverter::writeH5AD(spe, file = "../inst/extdata/spe.h5ad", compression = "gzip")
```


# Summary

The downstream analysis of the spatial transcriptomics data depends on the biological questions being asked, `hoodscanR` and `scider` are designed to look at spatial transcriptomics data in different perspectives, therefore offering .


# Packages used 

This workflow depends on various packages from version `r BiocManager::version()` of the Bioconductor project, running on `r version$version.string` or higher. The complete list of the packages used for this workflow are shown below:

```{r session_info}
sessionInfo()
```

